---
title: "Best practices"
author: "Andrey A Shabalin"
date: "`r doc_date()`"
package: "`r pkg_ver('filematrix')`"
output:
  BiocStyle::html_document:
    toc: true # table of content true
vignette: >
  %\VignetteIndexEntry{2 Best practices}
  %\VignetteEngine{knitr::rmarkdown}
---

# Working with large matrices

The `filematrix` package can be used for matrices of any size.
The most convenient way of working with small and moderately sized matrices is to quickly save and load them via `fm.create.from.matrix` and `fm.load` respectively.

However, the main purpose of the `filematrix` package is to allow users to work with matrices many times larger than the amount of computer memory.
Such matrices can only be accessed by parts.

Let us setup a sufficiently large matrix for code examples below:

```{r setup, echo=FALSE}
  library(knitr)
  opts_knit$set(root.dir=tempdir())
```

```{r message=FALSE}
library(filematrix)
fm = fm.create(filenamebase = tempfile(), nrow = 5000, ncol = 10000, type = "double")
```

# Accessing all matrix elements

The fastest way to read or write all elements of a filematrix
is to work with columns sequentially, several columns at a time.
It is much faster than accessing a filematrix by rows.

The three examples below illustrate how this can be done for such tasks as
- filling filematrix with values, 
- calculating column sums, and
- calculating row sums.


## Filling in matrix values

Let us fill in the matrix with random values 512 columns at a time.
```{r}
step1 = 512
runto = ncol(fm)
nsteps = ceiling(runto/step1)
for( part in seq_len(nsteps) ) { # part = 1
 	fr = (part-1)*step1 + 1
 	to = min(part*step1, runto)
	cat( "Filling in columns", fr, "to", to, "\n")
 	fm[,fr:to] = runif(nrow(fm) * (to-fr+1))
}
rm(part, step1, runto, nsteps, fr, to)
```

## Calculating column sums

Let us calculate column sums of the filematrix, 256 columns at a time.
```{r}
fmcolsums = double(ncol(fm))

step1 = 512
runto = ncol(fm)
nsteps = ceiling(runto/step1)
for( part in seq_len(nsteps) ) { # part = 1
 	fr = (part-1)*step1 + 1
 	to = min(part*step1, runto)
	cat( "Calculating column sums, processing columns", fr, "to", to, "\n")
 	fmcolsums[fr:to] = colSums(fm[,fr:to])
}
rm(part, step1, runto, nsteps, fr, to)

cat("Sums of first and last columns are", fmcolsums[1], "and", tail(fmcolsums,1), "\n")
```

## Calculating row sums

Let us calculate column sums of the filematrix, 256 columns at a time.
```{r}
fmrowsums = double(nrow(fm))

step1 = 512
runto = ncol(fm)
nsteps = ceiling(runto/step1)
for( part in seq_len(nsteps) ) { # part = 1
 	fr = (part-1)*step1 + 1
 	to = min(part*step1, runto)
	cat( "Calculating row sums, processing columns", fr, "to", to, "\n")
 	fmrowsums = fmrowsums + rowSums(fm[,fr:to])
}
rm(part, step1, runto, nsteps, fr, to)

cat("Sums of first and last rows are", fmrowsums[1], "and", tail(fmrowsums,1), "\n")
```

# Removing filematrix after use

```{r}
closeAndDeleteFiles(fm)
```
