\section{Other 3D tools in \RR\label{tools}}
At the time of writing \sdd , the function \code{persp()} in the base
package of \RR\ for three dimensional surface plots was available, but
there was no way to generate 3D scatter plots in \RR\ itself.

The data visualization system \emph{xgobi} (\cite{swayne98}) provides
interactive visualization of multidimensional data, e.g. brush and spin,
higher-dimensional rotation, grand tour, etc.  The \RR\ package
\emph{xgobi} (\cite{swayne91}; we have to distinguish the visualization
system and the package) provides an Interface to \emph{xgobi} and launches
a \emph{xgobi} process appropriately.
%%
\emph{ggobi}\footnote{\url{http://www.ggobi.org}}
(\cite{swayne02}) is the next edition of \emph{xgobi}.

Analogously to \emph{xgobi} a \RR\ package \emph{Rggobi} (\cite{temple01})
exists in the \emph{Omegahat} project\footnote{\url{http://www.omegahat.org}}
(\cite{temple00}) that allows one to embed \emph{ggobi} within
\RR\ and to both set and query the \emph{ggobi} contents.  All in all,
\emph{ggobi} can be loaded dynamically into \RR\ (as well as into other
software products, in principle), and \RR\ into \emph{ggobi}.  This
provides interactive, direct manipulation, linked, high-dimensional
graphics within \RR .

The package \emph{rgl}\footnote{\url{http://wsopuppenkiste.wiso.uni-goettingen.de/~dadler/rgl/}}
by \cite{AdlerNenadic2003} is a portable \RR\ programing interface to \emph{OpenGL}.
Its features include, e.g., interactive viewpoint navigation,
automatic data focus, up to 8 light sources, alpha-blending
(transparency), and environmental effects like fogging.
The package \emph{djmrgl}\footnote{This package was
formerly called \emph{rgl}, similar to the other mentioned package
by Daniel Adler and Oleg Nenadi\'{c}.
\emph{djmrgl} is only available for the Windows operating system.}
(\cite{murdoch}) also provides an \RR\ interface to \emph{OpenGL}.
A huge collection of useful functions to generate,
manipulate and interactively rotate 3D objects is available.
Efforts are under way to merge these two packages.

The function \code{cloud} in the lattice package is a 3D scatter plot function that works in the
\emph{lattice} (\cite{sarkar02}) (and \emph{grid} (\cite{murrell01}))
environment of \RR .
\emph{Lattice} is an implementation of \emph{Trellis Graphics}, which is a
framework for data visualization developed at the Bell Labs by
\citeN{becker96}, extending ideas presented in \citeN{cleveland}.

