---
title: "Formula Interface for ggplot2"
author: "Daniel Kaplan and Randall Pruim"
date: "June 17, 2017"
output:  rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteIndexEntry{Formula Interface for ggplot2}
---

```{r include = FALSE}
library(ggformula)
library(dplyr)
library(ggplot2)
library(mosaicData)
knitr::opts_chunk$set(
  fig.show = "hold",
  out.width = "45%"
)
theme_set(theme_light())
```


# Formula-driven graphics


There are several excellent graphics packages provided for R. The
`ggformula` package currently builds on one of them, `ggplot2`, but
provides a very different user interface for creating plots. The interface
is based on formulas (much like the `lattice` interface) and the use of 
the chaining operator (`%>%`) to build more complex graphics from simpler 
components.


Experts, of course, will want to use the underlying `ggplot2` package directly
in order to maximize expressiveness and minimize the "bureaucracy" induced by
intervening layers of software. The `ggformula` graphics are suited
for 

  * beginners, who don't require the full range of `ggplot2` capabilities, but
  want to get started quickly,
  
  * those familiar with `lattice` graphics, but wanting to be
  able to easily create multilayered plots,
  
  * those who prefer a formula interface, perhaps because it is 
  familiar from use with function like `lm()` or from use of 
  the `mosaic` package for numerical summaries.
  

## The basic formula template

The basic template for creating a plot with `ggformula` is

```{r, eval = FALSE}
gf_plottype(formula, data = mydata)
```

where 

  * `plottype` describes the type of plot (layer) desired (points, lines, a histogram,
etc., etc.), 

  * `mydata` is a data frame containing the variables used in the plot, and 
  
  * `formula` describes how/where those variables are used.

For example, in a bivariate plot, `formula` will take the form `y ~ x`, where `y` is the 
name of a variable to be plotted on the y-axis and `x` is the name of a variable to 
be plotted on the x-axis.  (It is also possible to use expressions that can be evaluated
using variables in the data frame as well.)

Here is a simple example:
```{r}
library(ggformula)
gf_point(mpg ~ hp, data = mtcars)
```



## Selecting the glyph type

The "kind of graphic" is specified by the name of the graphics function. All of
the `ggformula` data graphics functions have names starting with
`gf_`, which is intended to remind the user that they are formula-based
interfaces to `ggplot2`: `g` for `ggplot2` and `f` for "formula." 
Commonly used functions include

- `gf_point()` for scatter plots
- `gf_line()` for line plots (connecting dots in a scatter plot)
- `gf_density()` or `gf_dens()` or `gf_histogram()` or `gf_freqpoly()` to display distributions of a quantitative variable
- `gf_boxplot()` or `gf_violin()` for comparing distributions side-by-side
- `gf_counts()` for bar-graph style depictions of counts.
- `gf_bar()` for more general bar-graph style graphics

The function names generally match the corresponding function name from `ggplot2`,
although `gf_counts()` is a simplified special case, and `gf_dens()` is an 
alternative to `gf_density()` that displays the density plot slightly differently
than the default in `ggplot2`.

Each of the `gf_` functions can create the coordinate axes and fill it in one
operation. (In `ggplot2` nomenclature, `gf_` functions create a frame and add a
geom layer, all in one operation.)  This is what happens for the first
`gf_` function in a chain.  For subsequent `gf_` functions, new layers are added,
each one "on top of" the previous layers.

## Attributes

Each of the marks in the plot is a *glyph*. Every glyph has graphical *attributes*
(called aesthetics in `ggplot2`) that tell where and how to draw the glyph. 
In the above plot, the obvious attributes are x- and y-position:  
We've told R to put `mpg` along the y-axis and `hp` along the x-asis, as is clear
from the plot. 
But each point also has other attributes, including color, shape, size, stroke, fill, 
and alpha (transparency).  We didn't specify
those in our example, so `gf_point()` uses some default values for those --
in this case smallish black filled-in circles.

### Specifying attributes

In the `gf_` functions, you specify the non-position graphical attributes using 
an extension of the basic formula.  Attributes can be **set** to a constant value
(e.g, set the color to "blue"; set the size to 2) 
or they can be **mapped** to a variable in the data or some 
expression involving the variables 
(e.g., map the color to `sex`, so sex determines the color groupings)

Attributes are set or mapped using additional arguments.

 * adding an argument of the form `attribute = value` **sets** `attribute` to `value`.
 * adding an argument of the form `attribute = ~ expression` **maps** `attribute` to `expression`
 
where `attribute` is one of `color`, `shape`, etc., `value` is a constant 
(e.g. `"red"` or `0.5`, as appropriate), and `expression`
may be some more general expression that can be computed using the variables in `data` 
(although often is is better to create a new variable in the data and to
use that variable instead of an on-the-fly calculation within the plot).


The following plot, for instance, 

 * We use `cyl` to determine the color and `carb` to  determine the size of each
 dot.  Color and size are **mapped** to `cyl` and `carb`. 
 A legend is provided to show us how the mapping is being done.
 (Later, we can use scales to control precisely how the mapping is done -- 
 which colors and sizes are used to represent which values of `cyl` and `carb`.)  
 
 * We also **set** the transparency to 50%.  The gives the same value of `alpha` to
 all glyphs in this layer.
 
```{r}
gf_point(mpg ~ hp, color = ~ cyl, size = ~carb, alpha = 0.50, data = mtcars) 
```

Here is the same plot where mapping and setting are indicated 
explicitly.
```{r}
# set alpha using a function argument instead of in the formula
gf_point(mpg ~ hp, color = ~cyl, size = ~carb, alpha = 0.50, data = mtcars) 
```

### On-the-fly calculations

`ggformula` allows for on-the-fly calculations of attributes, although the default labeling 
of the plot is often better if we create a new variable in our data frame.  In the 
examples below, since there are only three values for `carb`, it is easier to read the 
graph if we tell R to treat `cyl` as a categorical variable by converting to a factor (or to 
a string).  Except for the labeling of the legend, these two plots are the same.

```{r}
gf_point(mpg ~ hp,  color = ~factor(cyl), size = ~carb, alpha = 0.75, data = mtcars)
gf_point(mpg ~ hp,  color = ~cylinders, size = ~carb, alpha = 0.75, 
         data = mtcars %>% mutate(cylinders = factor(cyl)))
```


## "One-variable" plots

For some plots, we only have to specify the x-position because the y-position is calculated
from the x-values.  Histograms, densityplots, and frequency polygons are examples.
To illustrate, we'll use density plots, but the same ideas apply to 
`gf_histogram()`, and `gf_freqpolygon()` as well. 
*Note that in the one-variable density graphics, the variable whose density is to be calculated goes to the right of the tilde, in the position reserved for the x-axis variable.*

```{r fig.show = "hold", out.width = "30%", warning=FALSE}
data(Runners, package = "statisticalModeling")
Runners <- Runners %>% filter( ! is.na(net))
gf_density( ~ net, data = Runners)
gf_density( ~ net,  fill = ~sex,  alpha = 0.5, data = Runners)
# gf_dens() is similar, but there is no line at bottom/sides, and it is not "fillable"
gf_dens( ~ net, color = ~sex, alpha = 0.7, data = Runners)    
```

Several of the plotting functions include additional arguments that do not modify
attributes of individual glyphs but control some other aspect of the plot.  In this
case, `adjust` can be used to increase or decrease the amount of smoothing.

```{r}
# less smoothing
gf_dens( ~ net, color = ~sex, alpha = 0.7, data = Runners, adjust = 0.25)  
# more smoothing
gf_dens( ~ net, color = ~sex, alpha = 0.7, data = Runners, adjust = 4)     
```

## Position

When the `fill` or `color` or `group` aesthetics are mapped to a variable, the
default behavior is to lay the group-wise densities on top of one another. Other
behavior is also available by using `position` in the formula. Using the value
`"stack"` causes the densities to be laid one on top of another, so that the
overall height of the stack is the density across all groups. The value `"fill"`
produces a conditional probability graphic.
```{r fig.show = "hold", warning=FALSE}
gf_density( ~ net, fill = ~sex, color = NA, position = "stack", data = Runners)
gf_density( ~ net, fill = ~sex, color = NA, position = "fill", data = Runners, adjust = 2)
```

Similar commands can be constructed with `gf_histogram()` and `gf_freqpoly()`,
but note that `color`, not `fill`, is the active attribute for frequency
polygons.  It's also rarely good to overlay
histograms on top of one another -- better to use a density plot or a frequency 
polygon for that application.

## More 2-variable plots

### Using jitter and transparency to handle overlapping cases

Sometimes you have so many points in a scatter plot that they obscure one another. 
The `ggplot2` system provides two easy ways to deal with this: translucency and jittering.

Use `alpha = 0.5` to make the points semi-translucent. If there are many points
overlapping at one point, a much smaller value of alpha, say `alpha = 0.01`. 
We've already seen this above.

Using `gf_jitter()` in place of `gf_point()` will move the plotted points to
reduce overlap.  Jitter and transparency can be used together as well.

```{r}
gf_point(age ~ sex, alpha = 0.05, data = Runners)
gf_jitter(age ~ sex, alpha = 0.05, data = Runners)
```

### Box and Whisker plots

Box and whisker plots show the distribution of a quantitative variable as a
function of a categorical variable. The formula used in `gf_boxplot()` should
have the quantitative variable to the left of the tilde.  (To make horizontal 
boxplots using `ggplot2` you have to make vertical boxplots and then flip the 
coordinates with `coord_flip()`.)

```{r fig.show = "hold", warning = FALSE}
gf_boxplot(net ~ sex, color = "red", data = Runners)
gf_boxplot(net ~ sex, color = ~start_position, data = Runners)
```

This plot may surprise you.

```{r}
gf_boxplot(net ~ year, data = Runners)
```

This plot is placing a single box and whisker plot at the mean value of `year`.
The warning message suggests that we need to tell R how to form the groups
when using a quantitative variable for `x`.
It suggests using the `group` aesthetic, and sometimes, this is 
just what we want.

```{r}
gf_boxplot(net ~ year, group = ~ year, data = Runners)
```

But often, is is better to convert a discrete quantitative variable used for grouping into
a categorical variable (a factor or character vector). This can be done in several ways:
```{r}
# add a new variable to the data
Runners$the_year <- as.character(Runners$year)               # in base R
Runners <- Runners %>% mutate(the_year = as.character(year)) # in dplyr
gf_boxplot(net ~ the_year, color = ~sex, data = Runners)
# or do it on the fly
gf_boxplot(net ~ factor(year), color = ~sex, data = Runners)
```

### 2-dimensional density plots

Two-dimensional plots of density also have both a left and right component to the formula.

```{r fig.show = "hold"}
gf_density_2d(net ~ age, data = Runners)
gf_hex(net ~ age, data = Runners)
```

### Paths and lines

The `ggplot2` system offers two ways to connect points. `gf_line()` ignores the order of the points in the data, and draws the line going from left to right. `gf_path()` goes from point to point according to the order in the data. Both forms can use a `color` or `group` aesthetic as a flag to draw groupwise lines.

```{r fig.show = "hold", out.width = "30%"}
# create a categorical variable
mtcars <- mtcars %>% mutate(n_cylinders = as.character(cyl)) 
gf_line(mpg ~ hp, data = mtcars)
gf_path(mpg ~ hp, data = mtcars)
gf_line(mpg ~ hp, color = ~n_cylinders, data = mtcars)
```

The above are examples of *bad plots*. The viewer is unnecessarily distracted by
the zigs and zags in the connecting lines. It would be better to use
`gf_point()` here, but then you wouldn't see how `gf_line()` and `gf_path()`
work!

Here's a more useful example.  We begin with a scatter plot showing the number of 
live births in the US for each day of 1978.

```{r}
library(mosaicData)
gf_point(births ~ date, data = Births78)
```

Can this interesting pattern be explained by a weekday/weekend effect?  We could use 
color to show the days of the week.

```{r}
library(mosaicData)
gf_point(births ~ date, color = ~wday, data = Births78)
```

Converting to a line plot highlights the pattern and makes it easier to spot the unusual days.

```{r}
gf_line(births ~ date, color = ~wday, data = Births78)
```



## Plots with more than 2 positional attributes

Some layers require more than two attributes.  Typically this happens when the glyphs
of a layer are complex objects that could have been made using multiple layers, but 
belong together conceptually.  Examples include

  * `gf_pointrange()` -- plots a dot flanked by a line segment.  
  * `gf_linrange()` -- like `gf_pointrange()` but without the point.
  * `gf_errorbar()` -- vertical error bars.
  * `gf_errorbarh()` -- horizontal error bars.
  * `gf_ribbon()` -- a band between a line above and line below.
  
Often these are used to depict some sort of estimate of uncertainty in a
measurement or a prediction, but they can be used to represent any data of the
correct form.  Here we will use `gf_linerange()` and `gf_ribbon()` 
to indicate the high and low temperatures in New York for the first few months of 2013.

```{r}
require(weatherData)
Temps <- NewYork2013 %>%
  mutate(date = lubridate::date(Time),
         month = lubridate::month(Time)) %>% 
  filter(month <= 4) %>%
  group_by(date) %>%
  summarise(
    hi = max(Temperature, na.rm = TRUE),
    lo = min(Temperature, na.rm = TRUE)
  )   
gf_linerange(lo + hi  ~ date, color = ~hi, data = Temps)
gf_ribbon(lo + hi ~ date, data = Temps, color = "navy", alpha = 0.3)
```

## Positions and Stats

### Positions

`position_dodge()`, 
`position_jitter()`, and
`position_jitterdodge()` can be used to adjust the positions at which 
glyphs are placed.  Jittering adds some random noise and can be useful when
many observations have the same value.  Dodging moves groups of glyphs a fixed
difference to make it easier to distinguish the groups.

```{r}
gf_point(length ~ sex, color = ~ domhand, data = KidsFeet,
         position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.4))
```

### Stats

A **stat** is a transformation that is applied to the data before 
a plot is generated.  Several of the plots we have seen have made use 
of stats.

 * `gf_histogram()` uses `stat_bin()` to bin the data and count the 
 number of observations in each bin.  It is equivalent to
 
```{r}
gf_bar( ~ age, data = HELPrct, stat = "bin")
```
  * `gf_boxplot()` uses `stat_boxplot()` to compute the five-number summary on which
  the boxplot is based.
  
  * `gf_violin()`, `gf_density()`, and `gf_density2d()` use stats to compute a 
  sequence of points along an estimated density.  It is these points, and the not
  the raw data that are used to create the plot.
 
Mostly, the stats selected by default are just the ones you need. But sometimes
it is useful to select a different stat.  The `stat_summary()` and 
`stat_summary_bin()` stats are particularly useful in this respect.
These stats use a function to aggregate over unique values of `x` or over 
bins of `x` values and save the user needing to do that data transformation
manually. 

The default function applied in each group is `mean_se()`, which computes 
the mean (and the mean plus and minus one standard error)
This makes it simple to create an "interaction plot".

```{r}
gf_jitter(length ~ sex, color = ~domhand, data = KidsFeet,
          width = 0.1, height = 0) %>%
  gf_line(length ~ sex, color = ~domhand, data = KidsFeet,
          group = ~domhand, stat="summary")
```

The other two values computed by `mean_se()` are available as
`..ymin..` and `..ymax..`.

```{r}
gf_jitter(length ~ sex, color = ~domhand, data = KidsFeet,
          width = 0.1, height = 0, alpha = 0.3) %>%
  gf_pointrange(length + ..ymin.. + ..ymax.. ~ sex, 
                color = ~domhand, data = KidsFeet, 
                group = ~domhand, stat="summary")
```

Custom functions can be used by defining `fun.y`, `fun.ymin`, and `fun.ymax`,
or a single function `fun.data` that returns a data frame with variables 
named `y`, `ymin`, and `ymax`.


```{r}
gf_point(length ~ sex, color = ~domhand, data = KidsFeet,
          width = 0.1, height = 0, alpha = 0.5,
          position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.3)) %>%
  gf_pointrange(length + ..ymin.. + ..ymax.. ~ sex, 
                color = ~domhand, data = KidsFeet, 
                group = ~domhand, stat="summary",
                fun.y = median, fun.ymin = min, fun.ymax = max,
                position = position_dodge(width = 0.6))
```

## Chaining to create complex plots

### Multiple layers 

Often it is useful to overlay multiple layers onto a single plot.  This can be done 
by chaining them with `%>%`, the "then" operator from `magrittr`.  The `data` 
argument can be omitted if the new layers uses the same data as the first layer 
in the chain. 

The following plot illustrates how histograms and frequency polygons are related.

```{r}
gf_histogram( ~ age, data = Runners, alpha = 0.3, fill = "navy") %>%
  gf_freqpoly( ~ age)
```

A 2-d density plot can be augmented with a scatterplot.

```{r}
gf_density_2d(net ~ age, data = Runners) %>%
  gf_point(net ~ age, alpha = 0.01) 
```


### Faceting

The `ggplot2` system allows you to make subplots --- called "facets" --- based on
the values of one or two categorical variables. This is done by chaining
with  `gf_facet_grid()` or `gf_facet_wrap()`. These functions use formulas 
to specify which variable(s) are to be used for faceting.

```{r fig.show = "hold", warning=FALSE}
gf_density_2d(net ~ age, data = Runners) %>% gf_facet_grid( ~ sex)
# the dot here is a bit strange, but required to make a valid formula
gf_density_2d(net ~ age, data = Runners) %>% gf_facet_grid( sex ~ .)
gf_density_2d(net ~ age, data = Runners) %>% gf_facet_wrap( ~ the_year)
gf_density_2d(net ~ age, data = Runners) %>% gf_facet_grid(start_position ~ sex)
```
One use of facets is to create side-by-side plots of different data sources.  
First we merge the data into a single data frame with a variable to indicate 
the original data source.  Then we use facets to separate
the data into the various sources for the display.
The redundant use of the `y` and `color` attributes for 
temperature makes it easier to compare across facets.

```{r}
require(weatherData)
Temps <- NewYork2013 %>% mutate(city = "NYC") %>%
  bind_rows(Mumbai2013 %>% mutate(city = "Mumbai")) %>%
  bind_rows(London2013 %>% mutate(city = "London")) %>%
  mutate(date = lubridate::date(Time),
         month = lubridate::month(Time)) %>% 
  group_by(city, date) %>%
  summarise(
    hi = max(Temperature, na.rm = TRUE),
    lo = min(Temperature, na.rm = TRUE),
    mid = (hi + lo)/2
  )   
gf_ribbon(lo + hi ~ date, data = Temps, alpha = 0.3) %>%
  gf_facet_grid(city ~ .)

gf_linerange(lo + hi ~ date, color = ~ mid, data = Temps) %>%
  gf_facet_grid(city ~ .) %>%
  gf_refine(scale_colour_gradientn(colors = rev(rainbow(5))))
```

### Global plot adjustments


There are a number of things we may want to do to the entire plot -- adjusting
labels, colors, fonts, etc.  `ggformula` provides wrappers to the `ggplot2` functions 
for this so that the chaining syntax can be used.

```{r}
gf_histogram( ~ age, data = Runners, alpha = 0.2, fill = "navy",
              binwidth = 5) %>%
  gf_freqpoly( ~ age, binwidth = 5) %>%
  gf_labs(x = "age (years)", title = "Age of runners") %>%
  gf_lims(x = c(20, 80)) %>%
  gf_theme(theme = theme_minimal())

gf_histogram( ~ age, data = Runners, alpha = 0.5, fill = "white",
              binwidth = 5) %>%
  gf_freqpoly( ~ age, color = "skyblue", binwidth = 5, size = 1.5) %>%
  gf_labs(x = "age (years)", title = "Age of runners") %>%
  gf_lims(x = c(20, 80)) %>%
  gf_theme(theme = theme_dark())
```


