\HeaderA{pmml.lm}{Generate PMML for an lm object}{pmml.lm}
\keyword{interface}{pmml.lm}
\begin{Description}\relax
Generate the PMML (Predictive Model Markup Language) representation of
an \pkg{lm} object. The PMML can then be imported into other systems
that accept PMML.
\end{Description}
\begin{Usage}
\begin{verbatim}
## S3 method for class 'lm':
pmml(model, model.name="Regression_model", app.name="Rattle/PMML",
     description="Regression Model", copyright=NULL, ...)
\end{verbatim}
\end{Usage}
\begin{Arguments}
\begin{ldescription}
\item[\code{model}] an lm object.
\item[\code{model.name}] a name to give to the model in the PMML.
\item[\code{app.name}] the name of the application that generated the PMML.
\item[\code{description}] a descriptive text for the header of the PMML.
\item[\code{copyright}] the copyright notice for the model.
\item[\code{...}] further arguments passed to or from other methods.
\end{ldescription}
\end{Arguments}
\begin{Details}\relax
The generated PMML can be imported into any PMML consuming
application, such as Zementis' ADAPA.

Currently, the resultant PMML document will not encode interaction
terms.

Only numeric regression is supported currently. Generalised linear
models (logistic regression) are not yet supported.
\end{Details}
\begin{Author}\relax
\email{rguha@indiana.edu}
\end{Author}
\begin{References}\relax
Package home page: \url{http://rattle.togaware.com}

PMML home page: \url{http://www.dmg.org}

Zementis' useful PMML convert: \url{http://www.zementis.com/pmml_converters.htm}
\end{References}
\begin{SeeAlso}\relax
\code{\LinkA{pmml}{pmml}}.
\end{SeeAlso}

