\HeaderA{pmml.nnet}{Generate PMML for a nnet object}{pmml.nnet}
\keyword{interface}{pmml.nnet}
\begin{Description}\relax
Generate the PMML representation for a nnet object (Neural Network). 
The nnet object is converted into a PMML representation. The PMML 
can then be imported into other systems that accept PMML. 
With this code, a PMML representation can be obtained for Neural
Networks implementing classification (multi-class and binary) as well
as regression.
\end{Description}
\begin{Usage}
\begin{verbatim}
## S3 method for class 'nnet':
pmml(model, model.name="NeuralNet_model", app.name="Rattle/PMML",
     description="Neural Network PMML Model", copyright=NULL, ...)
\end{verbatim}
\end{Usage}
\begin{Arguments}
\begin{ldescription}
\item[\code{model}] a nnet object.
\item[\code{model.name}] a name to give to the model in the PMML.
\item[\code{app.name}] the name of the application that generated the PMML.
\item[\code{description}] a descriptive text for the header of the PMML.
\item[\code{copyright}] the copyright notice for the model.
\item[\code{...}] further arguments passed to or from other methods.
\end{ldescription}
\end{Arguments}
\begin{Details}\relax
The generated PMML can be imported into any PMML consuming
application that recognizes PMML 3.2. An example is ADAPA. 
ADAPA (Adaptive Decision and Predictive Analytics) is a lightweight 
decision engine that offers at its core batch and real-time scoring 
of predictive models as well as fast execution of business rules. 
ADAPA supports an extensive collection of PMML elements, including the 
following predictive techniques: 1) Neural Networks (Backprogagation 
and Neural Gas); 2) Support Vector Machines; 3) Linear and Logistic 
Regression as well as all general regression PMML models: a) Multinomial 
Logistic; b) General Linear; 3) Ordinal Multinomial; 4) Simple Regression; 
and 5) Generalized Linear Model. ADAPA  provides a reliable and fast 
way to manage, deploy, and execute a multitude of models and decision 
strategies.
\end{Details}
\begin{Author}\relax
\email{info@zementis.com}
\end{Author}
\begin{References}\relax
Package home page: \url{http://rattle.togaware.com}

PMML home page: \url{http://www.dmg.org}

Zementis' useful PMML convert: \url{http://www.zementis.com/pmml_converters.htm}

ADAPA home page: \url{http://www.zementis.com/adapa.htm}
\end{References}
\begin{SeeAlso}\relax
\code{\LinkA{pmml}{pmml}}.
\code{\LinkA{nnet}{nnet}}.
\end{SeeAlso}

